---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# hashtagassignment

<!-- badges: start -->
<!-- badges: end -->

In 10x and other single cell sequencing technologies, antibody barcodes are
routinely added to cells before sequencing in order to multiplex multiple
experiments in a single sequencing run.  Existing hash tag assignment strategies
(e.g. HTODemux) implicitly assume total hash tag read counts to be identical
between cells, but this assumption is false. We observed a strong correlation
between the total hash tag read counts and the number of unique UMIs. The
strategy presented here therefore doesn’t use hashtag thresholds that are shared
across cells, but rather determines the most likely hashtag on a per cell basis.
It removes doublets and low confident cells by thresholding on i) the log2 ratio
of the maximal hash tag’s read count with the second best hash tag ii) the
‘evenness’ (normalized entropy) of hash tag counts and iii) the total number of
reads assigned to hash tags. The second metric will be low if one hash tag
clearly is the winner and higher if reads are more evenly distributed across
hash tags. 

## Installation

```{r, eval = FALSE}
if (!require('hashtagassignment')) {
  # install.packages("devtools")
  # devtools::install("~/libs/hashtagassignment")
  devtools::install_github("slagtermaarten/hashtagassignment")
}
library(hashtagassignment)
```

```{r, include = FALSE, warning=FALSE}
knitr::opts_knit$set(keep_md = TRUE, cache = TRUE)
```

## Example usage

```{r, warning=FALSE}
library(dplyr)
data_dir <- '/DATA/users/m.slagter/MirjamHoekstra/raw_exp_5310'
hashtag_counts <- extract_hashtags_from_cellranger(data_dir = data_dir)

## Define thresholds
## Cells need the winning hash tag to be at least 2^2 (=4) times larger than the
## second best hash hag
fd_thresh = 2
## Cells can have a hashtag evenness of at most .5
evenness_thresh = .5
## Minimal amount of hash tag reads cells need to have
read_thresh = 100
stats <- compute_hashtag_stats(hashtag_counts, 
                               fd_thresh = fd_thresh,
                               evenness_thresh = evenness_thresh, 
                               read_thresh = read_thresh)

## Filter for cells that do meet all of the criteria 
if (exists('seurat_object')) {
  # seurat_object <- load_seurat('exp5310')
  # rownames(seurat_object@meta.data) <- 
  #   names(seurat_object@active.ident) <-
  #   gsub('5310_', '', rownames(seurat_object@meta.data))
  filtered_seurat_object <- filter_seurat_stats(seurat_object, stats)
}
```

Select cells to serve as examples for different levels of evenness ([0, .1, .2,
...]). Notice how the distributions get more fuzzy as evenness increases.

```{r, results='asis', warning=FALSE}
hashtag_evenness_examples <- purrr::map_dfr(seq(0, 1, by = .1), function(e) {
  idx <- order((stats$hashtag_evenness - e)^2)[1:2]
  stats %>%
    { .[idx, ] } %>%
    dplyr::select(sample_id, 
                  matches('HTO|human_Hashtag'), hashtag_evenness, fd_cc)
  }) %>% 
  unique %>% 
  dplyr::rename_with(.fn = function(x) gsub('(HTO\\d+)_.*', '\\1', x)) %>%
  dplyr::rename_with(.fn = function(x) gsub('human_Hashtag', 'HTO', x)) %>%
  dplyr::mutate(hashtag_evenness = round(hashtag_evenness, 2)) %>%
  dplyr::mutate(fd_cc = round(fd_cc, 2)) %>%
  { . }

hashtag_evenness_examples$selected <-
  with(hashtag_evenness_examples, 
       is.na(fd_cc) | fd_cc >= fd_thresh,
       hashtag_evenness <= evenness_thresh)
knitr::kable(hashtag_evenness_examples)
```

Create panel of plots to assess what kind and how many cells will be filtered
out

```{r, warning=FALSE}
library(gridExtra)
library(ggplot2)
library(patchwork)

p1 <- stats$total_hashtag_reads %>% log10 %>% 
  qplot(bins = 100) + xlab('log10(# hashtag reads)')

p2 <- ggplot(mapping = aes(x = hashtag_evenness, y = fd_cc),
             data = stats) +
  geom_hex(bins = 150) +
  xlab('Hashtag evennness') +
  geom_hline(yintercept = fd_thresh) +
  geom_vline(xintercept = evenness_thresh) +
  ylab('log2 fold difference\n of winning\n with closest contender') +
  ggplot2::theme(legend.position = c(.95, .95), legend.justification = c(1, 1))

p4 <- ggplot(mapping = aes(x = hashtag_evenness), data = stats) +
  geom_histogram(bins = 100) +
  xlab('Hashtag evennness') +
  ylab('Number of cells')

p5 <- stats %>%
  dplyr::group_by(fd_crit, evenness_crit, read_crit) %>%
  dplyr::count() %>%
  dplyr::mutate(frac = round(n / nrow(stats), 3)) %>%
  gridExtra::tableGrob(theme = gridExtra::ttheme_default(base_size = 6))

(p1 + p2) / (p4 + p5) 
```
